Quasi-universal acceptance of basic science in the US

### 2024-09-12 paper.tex
first version within word limit of 4k

Trust in science is related to many desirable outcomes, from acceptance of anthropogenic climate change (ref) or vaccination (ref) to following recommendations during COVID (Algan, Cohen, Davoine, Foucault, \& Stantcheva, 2021), which suggests that trust in science was the most important predictor of these behaviors).

@@

What does this apparent rejection of science actually entail? Do people who say they do not trust science much, or who believe in conspiracy theories at odds with well-established science, reject most of science? Or, on the contrary, do they object to a few specific facets of science, while still accepting the overwhelming majority of basic science? Answering this question has theoretical implications: if some people reject science wholesale, then it's possible that their distrust of science causes their rejection of specific scientific theories, or facilitates their acceptance of conspiracy theories; by contrast, if everyone trusts most of science, then the mistrust of specific facets of science is more likely to be post-hoc, a rationalization of other beliefs or behaviors. The present question also has practical implications: many communication attempts leverage the scientific consensus (e.g.~on vaccination, climate change, etc., for review, see Van Stekelenburg, Schaap, Veling, Van 'T Riet, \& Buijzen, 2022; see also Većkalov et al., 2024). These attempts are more likely to be successful if everyone trusts basic science than if some people reject science wholesale.

This relative lack of trust in science could mean different things. It could mean that some people reject science wholesale, or at least to a significant extent. Or it could mean that they maintain a high degree of trust in most of science, and only question specific results that happen to contradict specific other beliefs they have (e.g.~creationism), or beliefs that are associated with disliked behaviors (e.g.~vaccination, acting for climate change).

We answer this question by looking at how much Americans accept basic scientific knowledge. We pay special attention to two groups of participants: people who say they don't trust science (much, or at all), and people who endorse science-related conspiracy theories (CTs), or have a CT mindset, since this has been linked with a low degree of trust in science, and that many CTs imply a mistrust of at least some science (vaccination, hydroxychloroquine, etc.).

We briefly review work on trust in science, and science knowledge, before turning to work on CT and trust in science.

\subsection{Trust in science}\label{trust-in-science}

Across the globe, most people say they trust science at least to some extent. In 2018, the Wellcome Global Monitor (WGM) surveyed of over 140'000 people in over 140 countries on trust in science (Wellcome Global Monitor, 2018). In 2020, during the first year of the Covid pandemic, a follow up survey was made in 113 countries, involving 119'000 participants (Wellcome Global Monitor, 2020). Across all countries, in 2018, 32\% of participants said they trust science ``a lot'' (41\% in 2020), 45\% trust ``some'' science (39\% in 2020), and only 13\% percent trust science ``not much/ not at all'' (13\% in 2020), while 10\% indicated ``don't know'' (7\% in 2020)\footnote{\url{https://wellcome.org/reports/wellcome-global-monitor-covid-19/2020\#gid=6acd&pid=0}\textgreater{}}.

Trust in science also appears to be relatively stable in the US (Funk \& Kennedy, 2020). However, there are some potentially worrying trends: Recently, trust in US has dropped (Brian \& Tyson, 2023), and some evidence suggests increasing polarization of trust in science along political lines (Gauchat, 2012). For example, the partisan divide on the question of whether climate change is a major threat to society has been growing in the last decade\footnote{\url{https://www.pewresearch.org/short-reads/2023/08/09/what-the-data-says-about-americans-views-of-climate-change/}} (Kennedy, 2023). A recent study suggests that by some measure, only about half of the US population believes in anthropogenic climate change, and that individuals do mostly not seem to change their mind on on the issue (Motta, 2021).

Trust in science is typically measured in three ways: either asking explicitly about general trust in science or scientists, by asking about science attitudes (e.g.~support for public funding of science), or asking questions about specific, typically contentious, science topics (e.g.~vaccines, climate change, GMOs). It is unclear to which extent these measures capture trust in basic science. To assess this, we can look at the literature on science knowledge.

{[}introduce SK as: that gives us a lower bound of how much people trust basic science{]}

\subsection{Science knowledge}\label{science-knowledge}

Much of the research on public understanding of science measures to which extent people have basic science knowledge (e.g.~``Does the Earth go around the Sun, or does the Sun go around the Earth?''). For us, these surveys provide a lower bound to how much people trust basic science: people who give the right answer (and not by chance) supposedly trust science on that question. Unfortunately, that lower bound has often been indeed pretty low: researchers have pointed out the deficit in basic science knowledge (Miller, 1998, 2004). For example, only 55\% in the US (in 2014) and only 46\% in the EU (in 2005) knew that antibiotics kill only bacteria and not viruses (Science Literacy et al., 2016). A question that remains is whether people do not provide the right answers because they are not familiar with the scientific consensus, or because they explicitly reject that consensus.

{[}To do: Make a descriptive overview of science knowledge across time/countries.{]}

\subsection{CT and TiS}\label{ct-and-tis}

Recently, the relationship of one group to trust in science (TiS) has been particularly scrutinized: conspiracy theorists (CTists), either defined as people who endorse specific conspiracy theories, or as people who have a conspiracy mindset (Rutjens \& Većkalov, 2022; Vranic, Hromatko, \& Tonković, 2022). CTists reject specific science knowledge (related to their CT), and they tend to trust science less on the whole. What remains unclear is to which extent CTists (dis)trust basic science.

@@

In a series of four pre-registered online studies (total n = 782), we asked US participants questions about well-established scientific facts. For each question, we asked participants what they thought the correct answer was (testing their knowledge of science), we informed them of the scientifically accepted correct answer, and asked them whether they accepted it (measuring their trust in basic science). We also measured participants' trust in science using standard measures, as well as their beliefs in various conspiracy theories and their tendency to engage in conspiratorial thinking. The four studies, including materials, hypotheses, and analyses, were pre-registered and all materials and data are accessible via the Open Science Framework (OSF)). The differences between the four studies are summarized presently, and the methods are detailed below.

@@

In Study 1, we used questions drawn from questionnaires of scientific knowledge (e.g.~``Are electrons smaller, larger, or the same size as atoms? {[}Smaller; Same size; Larger{]}''), supplemented by a `trick' question (``Where do trees mainly draw the materials with which they create their mass? {[}Earth; Water; Air{]}''; correct answer: Air). In Studies 2 and 3, this last question was removed. The scientific facts used in Studies 1 to 3 represent long-established and basic knowledge. In Study 4 we used more recent much less basic scientific discoveries (e.g.~``How many more glial cells are there in the brain in comparison with neurons, as established in 2016? {[}The same amount; Twice as many; Tenth as many{]}''; correct answer: The same amount).

@@

In Study 1, we simply told participants that they would be provided with the scientifically consensual answer. However, for participants to accept this answer, they must not only trust science, but also trust us (that we are presenting them with the actual scientifically consensual answer). To remove this issue, in Studies 2 to 3, we presented participants with a short explanation of the correct answer, as well as links to three sources per answer (e.g.~Wikipedia, National Geographic or NASA). In Study 4, as the topics were more complex, we did not provide an explanation, but we provided two sources per answer.

@@

For study 2, we restricted our main hypotheses about acceptance to cases in which participants initially provided a wrong answer. The rationale was to see if trust in science or conspiracy belief were associated with a change of mind towards the consensus. We provide results on these conditional correlations--for Study 2 and for all other studies--in the ESM\footnote{Only in study 4 do we find evidence that changing one's mind towards the scientific consensus is associated with (more) trust in science (study 1: r = 0.06, p = 0.387; study 2: r = 0.16, p = 0.051; study 3: r = 0.04, p = 0.619; study 4: r = 0.15, p = 0.037) and only in study 2 evidence that it is associated with (less) conspiracy beliefs and (less) conspiracy thinking (study 1: r = -0.14, p = 0.061; study 2: r = -0.22, p = 0.006; study 3: r = -0.04, p = 0.631; study 4: r = -0.05, p = 0.455).}. However, for the analysis presented here, we proceeded as preregistered for all other studies, by reporting simple, unconditional correlations between acceptance and trust in science, or, respectively, conspiracy belief.

@@

After providing their consent to participate in the study, participants were given an attention check ``While watching the television, have you ever had a fatal heart attack?'' {[}1-6; 1 = Never, 6 = Often{]}. All participants who did not answer ``1 = Never'' were excluded. Participants then read the following instructions:``We will ask you 10 questions about science. After each question, we will provide you with the scientifically consensual answer and ask whether you accept it.'' Next, participants answered a set of 10 basic science questions (in Study 1 they were randomly selected from a pool of 11 questions, as one `trick' question had been added) in random order. After each question, participants were presented with an answer reflecting the scientific consensus, and asked whether they accepted it. In Studies 2 and 3, participants additionally saw a short explanation, partly based on explanations generated by ChatGPT, and three links to authoritative sources supporting the answer. In Study 4, we provided only two links and no explanation. Participants then answered questions on conspiracy thinking, conspiracy beliefs, and trust in science.

@@

After removing failed attention checks, we were left with a total sample size of 782 (194 in Study 1, 6 failed attention checks; 190 in Study 2, 11 failed attention checks; 200 in Study 3, no failed attention checks; 198 in Study 4, 2 failed attention checks) participants in the US via prolific. Details and demographics can be found in the online supplemental material. While samples for Studies 1 and 2 were convenience samples, Studies 3 and 4 were conducted on a sample holding vaccine-skeptic beliefs. Prolific Academic allows to select participants based on their answers to a range of questions. We picked three of these questions and only recruited participants who met our criteria for each of them:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  ``Please describe your attitudes towards the COVID-19 (Coronavirus) vaccines: {[}For (I feel positively about the vaccines); Against (I feel negatively about the vaccines); Neutral (I don't have strong opinions either way); Prefer not to say''{]}. We selected particpants who answered''Against''.
\item
  ``Have you received a coronavirus (COVID-19) vaccination? {[}Yes (at least one dose); No; Prefer not to answer{]}''. We select only people who answered ``No''.
\item
  ``On a scale from 1-7, please rate to what extent you agree with the following statement: I believe that scheduled immunizations are safe for children. {[}1 (totally disagree); 2 (disagree); 3 (somewhat disagree); 4 (neither agree nor disagree); 5 (somewhat agree); 6 (agree); 7 (totally agree); rather not say{]}''. We select only people who answered `1', `2', or `3'.
\end{enumerate}

@@

Studies 1 to 3 used 10 facts drawn from widely used questionnaires about science knowledge (Allum, Sturgis, Tabourazi, \& Brunton-Smith, 2008; Durant, Evans, \& Thomas, 1989; Miller, 1998) sometimes referred to as the ``Oxford scale'' (Gauchat, 2011). A `trick' question was added in Study 1 and removed as its wording proved unclear. Study 4 used 10 more recent scientific discoveries. Table \ref{tab:knowledge} shows all questions and their answer options used in the four studies.

@@

\begin{longtable}[t]{>{\raggedleft\arraybackslash}p{2em}>{\raggedright\arraybackslash}p{22em}>{\raggedright\arraybackslash}p{22em}}
\caption{\label{tab:knowledge}Science knowledge items}\\
\toprule
 & Study 1-3 & Study 4\\
\midrule
1 & Do antibiotics kill viruses as well as bacteria? [Yes, both; No, only viruses; No, only bacteria] & For which disease is the drug bedaquiline, developed in 2007, a treatment? [Tetanus; Tuberculosis; Malaria]\\
2 & Are electrons smaller, larger, or the same size as atoms? [Smaller; Same size; Larger] & What is the maximum speed a proton can attain in the largest particle collider as to 2015? [90\% of the speed of light; 99\% of the speed of light; the speed of light]\\
3 & Have the continents on Earth been moving for millions of years or have they always been where they are now? [They have been moving; They have always been where they are now] & Kepler-452b is an exoplanet revolving around the star Kepler-452. How far away from the star is it, as established by astronomers in 2015? [97 million mi; 1,2 million mi; 1254 million mi]\\
4 & What decides whether a baby is a boy or a girl ? Is it the father's genes, the mother's genes, or both? [The mother's genes; the father's genes; both] & Using bomb-pusle dating with carbon 14, what is the age of the oldest known vertebrate, as established in 2016? [138 years; 205 years;  392 years]\\
5 & Do lasers work by focusing sound waves? [Yes; No] & How many more glials cells are there in the brain in comparison with neurons, as established in 2016? [The same amount; Twice as many; Tenth as many]\\
\addlinespace
6 & How long does it take for Earth to go around the sun: one day, one month, or one year? [One day; One month; One year] & As predicted by the general theory of relativity, how many times would the Earth keep orbiting if the Sun disappeared, as established in 2012? [47 seconds; 8 minutes;  2 hours]\\
7 & Are diamonds made of carbon? [Yes; No] & What is the electric charge of the Higgs Boson, as established in 2012? [1.602176634 x 10-19; 0; 3.2x10-19C]\\
8 & Which travels faster : light or sound? [Light; Sound] & What is the age of the oldest materials formed on Earth, as established in 2020? [Less than 4.6 Ga; Around 4.6 Ga; More than 4.6 Ga]\\
9 & Is common table salt made of calcium carbonate? [Yes; No] & With the best current cloning techniques, what is the average success rate when operated on mice, as of 2010? [2,7\%; 9,4\%; 17,2\%]\\
10 & Is water made of molecules containing one oxygen and two hydrogen atoms? [Yes; No] & What was the strength of the Earth magnetic field 3.7 billion years ago, as discovered this year? [15 microtesla; 30 microtesla; 45 microtesla]\\
\addlinespace
11 & *Where do trees mainly draw the materials with which they create their mass? [Earth; Water; Air] & \\
\bottomrule
\multicolumn{3}{l}{\rule{0pt}{1em}\textsuperscript{*} Only used in Study 1}\\
\end{longtable}
\endgroup{}

@@

We selected 10 science/health related conspiracy theories from the Belief in Conspiracy Theory Inventory (BCTI) (Pennycook, Binnendyk, \& Rand, 2022) (Table \ref{tab:conspiracy}). Participants were asked: ``Below is a list of events for which the official version has been disputed. For each event, we would like you to indicate to what extent you believe the cover-up version of events is true or false. {[}1-9; labels: 1 - completly false, 5 - unsure, 9 - completely true{]}''.

@@

We used the four-item conspiracy mentality questionnaire (CMQ) (Bruder, Haffke, Neave, Nouripanah, \& Imhoff, 2013) and the single item conspiracy beliefs scale (SICBS) (Lantian, Muller, Nurra, \& Douglas, 2016). Details and comparisons between the scales can be found in the OSM.

@@

@@

The main outcome of interest is acceptance of the scientifically consensual facts presented. Overall, acceptance was very high (aggregating across all studies: 95.1 \%; Study 1: 93 \%; Study 2: 98 \%; Study 3: 98 \%; Study 4: 91 \%). Note that this includes both participants who had previously correctly answered the knowledge question, and participants who changed their mind when presented with the scientific consensus. In Studies 3 \& 4, we gave participants a second chance in case they had initially rejected the consensus, which slightly increased acceptance rates in those studies (initial acceptance in Study 3: 96 \%; in Study 4: 86 \%).

As shown in Figure \ref{fig:summary-plot}, these very high rates of acceptance hold for: participants who they do not trust science at all (5.7 \% of participants, acceptance rate of 87.3 \%), participants who rank in the top two deciles of the conspiracy thinking scale (28.8 \% of participants, average acceptance rate of 93.2 \%), participants who consider as ``completely true'' (the maximum of the 9-point scale) conspiracy theories stating that the earth is flat (5.8 \% of participants, acceptance rate of 87.2 \%), or that climate change due to fossil emissions is a hoax (12 \% of participants, acceptance rate of 91.7 \%). This doesn't merely reflect science knowledge: aggregated across all studies--participants only correctly answered 65.8 \% of the questions (Study 1: 74 \%; Study 2: 79 \%; Study 3: 75 \%; Study 4: 36 \%) before they were provided with the scientifically consensual answer. Did participants who had initially gotten it wrong change their minds towards the scientific consensus? Yes. In most cases (study 1: 76.3 \%; study 2: 92.9 \%; study 3: 95.5 \%; study 4: 89.5 \%), participants readily accepted the scientific consensus after having initially given the wrong answer to a question. In very few cases, participants changed their mind away from the scientific consensus. These participants initially gave the correct response, but rejected the scientific consensus right after, thereby contradicting their own initial response (study 1: 1.6 \%; study 2: 0.5 \%; study 3: 0.7 \%; study 4: 5.7 \%).

\begin{verbatim}
## # A tibble: 1 x 2
##   mean_acceptance mean_knowledge
##             <dbl>          <dbl>
## 1           0.805          0.416
\end{verbatim}

Regarding H1, we find a consistent association between trust in science and acceptance of the scientific consensus (Studies 1: r = 0.27, p \textless{} .001; 2: r = 0.30, p \textless{} .001; 3: r = 0.20, p = 0.019; 4: r = 0.16, p = 0.029), but a less consistent relation between trust in science and science knowledge (Studies 1: r = 0.29, p \textless{} .001; 2: r = 0.28, p \textless{} .001; 3: r = 0.14, p = 0.100; study 4: r = 0.10, p = 0.144).

Regarding H2, the results are mixed both for the relation between conspiracy beliefs (measured as the average acceptance of all the conspiracy beliefs) and acceptance of the scientific consensus (Studies 1: r = -0.33, p \textless{} .001; 2: r = -0.37, p \textless{} .001; 3: r = -0.02, p = 0.788; 4: r = -0.05, p = 0.498) as well as science knowledge (Studies 1: r = -0.38, p \textless{} .001; study 2: r = -0.40, p \textless{} .001; 3: r = -0.16, p = 0.055; 4: r = -0.02, p = 0.742).

Why did participants reject the scientific consensus? We got a total of 364 answers (study 2: 35; study 3: 74; study 4: 255) from 167 (study 2: 25; study 3: 47; study 4: 95) different participants to the open-ended questions on why they had rejected the scientific consensus on a particular question. Based on the answers, we created five categories. Table \ref{tab:justifications} summarizes these answers by five categories. All answers can be read in the OSM.

@@

\begin{tabular}{llll}
\toprule
Category & \multicolumn{1}{c}{N (instances)} & \multicolumn{1}{c}{Share (instances)} & \multicolumn{1}{c}{N (unique participants)}\\
\midrule
Not convinced & 158.00 & 43.4\% & 70.00\\
No justification & 103.00 & 28.3\% & 51.00\\
Personal convictions & 43.00 & 11.8\% & 31.00\\
Mistake & 42.00 & 11.5\% & 34.00\\
Religious Beliefs & 18.00 & 4.9\% & 13.00\\
\bottomrule
\end{tabular}

Why did participants accept the scientific consensus? In Studies 3 and 4--the vaccine skeptic samples--we had asked participants about cases where they agreed with the scientific consensus. A total of 320 (Study 3: 122; Study 4: 198) participants answered this question. There were more participants saying they accepted the scientific consensus because they independently verified the fact (Study 3: \texttt{exp3\_descriptives\$acceptance\_by\_reason{[}{[}"independent\ verification"{]}{]}\$share}; Study 4: 47\% ), than there were participants saying it was because they trust scientists (Study 3: 41.8\%; Study 4: 36.4\%). We asked participants who said they verified independently to explain how they did it (answers can be found in the ESM). 10.7\% in study 3 and 16.7\% in study 4 answered with other ``other'' and gave an open-ended explanation (see ESM).

>>>>

Trust in science is related to many desirable outcomes, from acceptance of anthropogenic climate change (Cologna \& Siegrist, 2020) or vaccination (Lindholt, Jørgensen, Bor, \& Petersen, 2021; Sturgis, Brunton-Smith, \& Jackson, 2021) to following recommendations during COVID (Algan, Cohen, Davoine, Foucault, \& Stantcheva, 2021), which suggests that trust in science was the most important predictor of these behaviors).

@@

What does this apparent rejection of science actually entail? Do people who say they do not trust science, or who believe in conspiracy theories at odds with well-established science, reject most of science? Or, on the contrary, do they object to a few specific facets of science, while still accepting the overwhelming majority of basic science? Answering this question has theoretical implications: if some people reject science wholesale, then it's possible that their distrust of science causes their rejection of specific scientific theories, or facilitates their acceptance of conspiracy theories; by contrast, if everyone trusts most of science, then the mistrust of specific facets of science is more likely to be post-hoc, a rationalization of other beliefs or behaviors. The present question also has practical implications: many communication attempts leverage the scientific consensus (e.g.~on vaccination, climate change, etc., for review, see Van Stekelenburg, Schaap, Veling, Van 'T Riet, \& Buijzen, 2022; see also Većkalov et al., 2024). These attempts are more likely to be successful if everyone trusts basic science than if some people reject science wholesale.

Studies of science knowledge provide relevant evidence, as people who answer science knowledge questions correctly presumably trust scientists in this respect. The average level of science knowledge is quite low, even in countries with extensive science education. For example, only 55\% in the US (in 2014) and only 46\% in the EU (in 2005) knew that antibiotics kill only bacteria and not viruses (Science Literacy et al., 2016). However, these results only provide a lower bound on how much people trust basic science, as participants who do not provide the correct answer might accept it when it is pointed out to them.

@@

In a series of four pre-registered online studies (total n = 782), we asked US participants questions about well-established scientific facts. For each question, we asked participants what they thought the correct answer was (testing their knowledge of science), we informed them of the scientifically accepted answer, and asked them whether they accepted it (measuring their trust in basic science). We also measured participants' trust in science using standard measures, as well as their beliefs in various conspiracy theories and their tendency to engage in conspiratorial thinking. The four studies, including materials, hypotheses, and analyses, were pre-registered and all materials and data are accessible via the Open Science Framework (OSF). The differences between the four studies are summarized presently, and the methods are detailed below.

@@

In Study 1, we used questions drawn from questionnaires of scientific knowledge (e.g.~``Are electrons smaller, larger, or the same size as atoms? {[}Smaller; Same size; Larger{]}''), supplemented by a `trick' question (``Where do trees mainly draw the materials with which they create their mass? {[}Earth; Water; Air{]}''; correct answer: Air). In Studies 2 and 3, this last question was removed. The scientific facts used in Studies 1 to 3 represent long-established and basic knowledge. In Study 4 we used more recent much less basic scientific discoveries (e.g.~``What is the electric charge of the Higgs Boson, as established in 2012? {[}1.602176634 × 10-19; 0; 3.2×10-19C{]}''; correct answer: 0, i.e.~electrically neutral).

@@

In Study 1, we simply told participants that they would be provided with the scientifically consensual answer. However, for participants to accept this answer, they must not only trust science, but also trust that we are presenting them with the actual scientifically consensual answer. To remove this issue, in Studies 2 to 3, we presented participants with a short explanation of the correct answer, as well as links to three sources per answer (e.g.~Wikipedia, National Geographic or NASA). In Study 4, as the topics were more complex, we did not provide an explanation, but still provided two sources per answer.

@@

For Study 2, we restricted our main hypotheses about acceptance to cases in which participants initially provided a wrong answer. However, this meant the more participants had initially provided correct answers, the fewer opportunities they had for accepting correct answers. We provide results on these conditional correlations--for Study 2 and for all other studies--in the ESM\footnote{Only in Study 4 do we find evidence that changing one's mind towards the scientific consensus is associated with (more) trust in science (Studies 1: r = 0.06, p = 0.387; 2: r = 0.16, p = 0.051; 3: r = 0.04, p = 0.619; 4: r = 0.15, p = 0.037) and only in Study 2 evidence that it is associated with (less) conspiracy beliefs and (less) conspiracy thinking (Studies 1: r = -0.14, p = 0.061; 2: r = -0.22, p = 0.006; 3: r = -0.04, p = 0.631; 4: r = -0.05, p = 0.455).}. However, for the analysis presented here, we proceeded as preregistered for all other studies, by reporting unconditional correlations between acceptance and trust in science, or, respectively, conspiracy belief.

@@

After providing their consent to participate in the study, participants were given an attention check ``While watching the television, have you ever had a fatal heart attack?'' {[}1-6; 1 = Never, 6 = Often{]}. All participants who did not answer ``1 = Never'' were excluded. Participants then read the following instructions:``We will ask you 10 questions about science. After each question, we will provide you with the scientifically consensual answer and ask whether you accept it.'' Next, participants answered a set of 10 basic science questions in random order. After each question, participants were presented with an answer reflecting the scientific consensus, and asked whether they accepted it. In Studies 2 and 3, participants additionally saw a short explanation, partly based on explanations generated by ChatGPT, and three links to authoritative sources supporting the answer. In Study 4, we provided only two links and no explanation. Participants then answered questions on conspiracy thinking, conspiracy beliefs, and trust in science.

@@

After removing failed attention checks, we were left with a total sample size of 782 (194 in Study 1, 6 failed attention checks; 190 in Study 2, 11 failed attention checks; 200 in Study 3, no failed attention checks; 198 in Study 4, 2 failed attention checks) participants in the US via prolific. Details and demographics can be found in the online supplemental material. While samples for Studies 1 and 2 were convenience samples, Studies 3 and 4 were conducted on a sample holding vaccine-skeptic beliefs. Prolific Academic allows selecting participants based on their answers to a range of questions. We picked three of these questions and only recruited participants who met our criteria for each of them:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  ``Please describe your attitudes towards the COVID-19 (Coronavirus) vaccines: {[}For (I feel positively about the vaccines); Against (I feel negatively about the vaccines); Neutral (I don't have strong opinions either way); Prefer not to say''{]}. We selected pariticpants who answered''Against''.
\item
  ``Have you received a coronavirus (COVID-19) vaccination? {[}Yes (at least one dose); No; Prefer not to answer{]}''. We select only people who answered ``No''.
\item
  ``On a scale from 1-7, please rate to what extent you agree with the following statement: I believe that scheduled immunizations are safe for children. {[}1 (totally disagree); 2 (disagree); 3 (somewhat disagree); 4 (neither agree nor disagree); 5 (somewhat agree); 6 (agree); 7 (totally agree); rather not say{]}''. We select only people who answered ``1'', ``2'', or ``3''.
\end{enumerate}

@@

Studies 1 to 3 used 10 facts drawn from widely used questionnaires about science knowledge (Allum, Sturgis, Tabourazi, \& Brunton-Smith, 2008; Durant, Evans, \& Thomas, 1989; Miller, 1998) sometimes referred to as the ``Oxford scale'' (Gauchat, 2011). A `trick' question was added in Study 1 and removed as its wording proved unclear. Study 4 used 10 more recent scientific discoveries. Table \ref{tab:knowledge} shows all questions and their answer options.

@@

\begin{longtable}[t]{>{\raggedleft\arraybackslash}p{2em}>{\raggedright\arraybackslash}p{22em}>{\raggedright\arraybackslash}p{22em}}
\caption{\label{tab:knowledge}Science knowledge items}\\
\toprule
 & Study 1-3 & Study 4\\
\midrule
1 & Do antibiotics kill viruses as well as bacteria? [Yes, both; No, only viruses; No, only bacteria] & For which disease is the drug bedaquiline, developed in 2007, a treatment? [Tetanus; Tuberculosis; Malaria]\\
2 & Are electrons smaller, larger, or the same size as atoms? [Smaller; Same size; Larger] & What is the maximum speed a proton can attain in the largest particle collider as to 2015? [90\% of the speed of light; 99\% of the speed of light; the speed of light]\\
3 & Have the continents on Earth been moving for millions of years or have they always been where they are now? [They have been moving; They have always been where they are now] & Kepler-452b is an exoplanet revolving around the star Kepler-452. How far away from the star is it, as established by astronomers in 2015? [97 million mi; 1,2 million mi; 1254 million mi]\\
4 & What decides whether a baby is a boy or a girl ? Is it the father's genes, the mother's genes, or both? [The mother's genes; the father's genes; both] & Using bomb-pusle dating with carbon 14, what is the age of the oldest known vertebrate, as established in 2016? [138 years; 205 years;  392 years]\\
5 & Do lasers work by focusing sound waves? [Yes; No] & How many more glial cells are there in the brain in comparison with neurons, as established in 2016? [The same amount; Twice as many; Ten times as many]\\
\addlinespace
6 & How long does it take for Earth to go around the sun: one day, one month, or one year? [One day; One month; One year] & As predicted by the general theory of relativity, how many times would the Earth keep orbiting if the Sun disappeared, as established in 2012? [47 seconds; 8 minutes;  2 hours]\\
7 & Are diamonds made of carbon? [Yes; No] & What is the electric charge of the Higgs Boson, as established in 2012? [1.602176634 x 10-19; 0; 3.2x10-19C]\\
8 & Which travels faster : light or sound? [Light; Sound] & What is the age of the oldest materials formed on Earth, as established in 2020? [Less than 4.6 Ga; Around 4.6 Ga; More than 4.6 Ga]\\
9 & Is common table salt made of calcium carbonate? [Yes; No] & With the best current cloning techniques, what is the average success rate when operated on mice, as of 2010? [2,7\%; 9,4\%; 17,2\%]\\
10 & Is water made of molecules containing one oxygen and two hydrogen atoms? [Yes; No] & What was the strength of the Earth magnetic field 3.7 billion years ago, as discovered this year? [15 microtesla; 30 microtesla; 45 microtesla]\\
\addlinespace
11 & **Where do trees mainly draw the materials with which they create their mass? [Earth; Water; Air] & \\
\bottomrule
\multicolumn{3}{l}{\rule{0pt}{1em}\textsuperscript{*} Only used in Study 1}\\
\end{longtable}
\endgroup{}

@@

We selected 10 science/health related conspiracy theories from the Belief in Conspiracy Theory Inventory (BCTI) (Pennycook, Binnendyk, \& Rand, 2022) (Table \ref{tab:conspiracy}). Participants were asked: ``Below is a list of events for which the official version has been disputed. For each event, we would like you to indicate to what extent you believe the cover-up version of events is true or false. {[}1-9; labels: 1 - completely false, 5 - unsure, 9 - completely true{]}''.

@@

We used the four-item conspiracy mentality questionnaire (CMQ) (Bruder, Haffke, Neave, Nouripanah, \& Imhoff, 2013) and the single item conspiracy beliefs scale (SICBS) (Lantian, Muller, Nurra, \& Douglas, 2016). Details and comparisons between the scales can be found in the ESM.

@@

\subsubsection{Trust in science}\label{trust-in-science}

@@

The main outcome of interest is acceptance of the scientifically consensual facts presented. Overall, acceptance was very high (aggregating across all studies: 95.1 \%; Study 1: 93 \%; Study 2: 98 \%; Study 3: 98 \%; Study 4: 91 \%). Note that this includes both participants who had previously correctly answered the knowledge question, and participants who changed their mind when presented with the scientific consensus. In Studies 3 and 4, we gave participants a second chance in case they had initially rejected the consensus, which slightly increased acceptance rates in those studies (initial acceptance in Study 3: 96 \%; in Study 4: 86 \%).

As shown in Figure \ref{fig:summary-plot}, these very high rates of acceptance hold for: participants who do not trust science at all (4.2 \% of participants, acceptance rate of 87.3 \%), participants who rank in the top two deciles of the conspiracy thinking scale (34.3 \% of participants, acceptance rate of 93.2 \%), participants who consider as ``completely true'' (the maximum of the 9-point scale) conspiracy theories stating that the earth is flat (3.7 \% of participants, acceptance rate of 87.2 \%), or that climate change due to fossil emissions is a hoax (11.4 \% of participants, acceptance rate of 91.7 \%).

Participants in the lowest decile of acceptance still had an average acceptance rate of 67.4 \%. Even participants who considered as ``completely true'' that the earth is flat and who said they do ``not trust science at all'' (N = 3) had an average acceptance rate of 86.7 \%.

These high acceptance rates do not merely reflect science knowledge: participants only correctly answered 65.8 \% of the questions (Study 1: 74 \%; Study 2: 79 \%; Study 3: 75 \%; Study 4: 36 \%) before they were provided with the scientifically consensual answer. Even the lowest decile in science knowledge which, on average, answered correctly only on 20 \% of the questions, had an average acceptance rate of 91.7 \%.

Did participants who had initially provided a wrong answer change their minds towards the scientific consensus? Yes. In most cases (study 1: 76.3 \%; study 2: 92.9 \%; study 3: 95.5 \%; study 4: 89.5 \%), participants readily accepted the scientific consensus after having initially given the wrong answer to a question.

Regarding H1, we find a consistent association between trust in science and acceptance of the scientific consensus (Studies 1: r = 0.27, p \textless{} .001; 2: r = 0.30, p \textless{} .001; 3: r = 0.20, p = 0.019; 4: r = 0.16, p = 0.029), but a less consistent relation between trust in science and science knowledge (Studies 1: r = 0.29, p \textless{} .001; 2: r = 0.28, p \textless{} .001; 3: r = 0.14, p = 0.100; 4: r = 0.10, p = 0.144).

Regarding H2, the results are mixed for the relation of conspiracy beliefs (measured as the average acceptance of all the conspiracy beliefs) with both acceptance of the scientific consensus (Studies 1: r = -0.33, p \textless{} .001; 2: r = -0.37, p \textless{} .001; 3: r = -0.02, p = 0.788; 4: r = -0.05, p = 0.498) and science knowledge (Studies 1: r = -0.38, p \textless{} .001; 2: r = -0.40, p \textless{} .001; 3: r = -0.16, p = 0.055; 4: r = -0.02, p = 0.742).

Why did participants reject the scientific consensus? We collected a total of 364 answers (study 2: 35; study 3: 74; study 4: 255) from 167 (study 2: 25; study 3: 47; study 4: 95) different participants to the open-ended questions on why they had rejected the scientific consensus on a particular question. Based on the answers, we created five categories. Table \ref{tab:justifications} summarizes these answers by five categories. All answers can be read in the ESM.

@@

\begin{tabular}{llll}
\toprule
Category & \multicolumn{1}{c}{N (instances)} & \multicolumn{1}{c}{Share (instances)} & \multicolumn{1}{c}{N (unique participants)}\\
\midrule
Not convinced & 158 & 43.4\% & 70\\
No justification & 103 & 28.3\% & 51\\
Personal convictions & 43 & 11.8\% & 31\\
Mistake & 42 & 11.5\% & 34\\
Religious Beliefs & 18 & 4.9\% & 13\\
\bottomrule
\end{tabular}

Why did participants accept the scientific consensus? In Studies 3 and 4--the vaccine hesitant samples--we had asked participants about cases in which they agreed with the scientific consensus. A total of 320 (Study 3: 122; Study 4: 198) participants answered this question. There were more participants saying they accepted the scientific consensus because they independently verified the fact (Study 3: 47.5\%; Study 4: 47\% ), than participants saying it was because they trust scientists (Study 3: 41.8\%; Study 4: 36.4\%)\footnote{10.7\% in Study 3 and 16.7\% in Study 4 answered with ``other'' and gave an open-ended explanation (see ESM).}. We asked participants who said they verified independently to explain how they did it (answers can be found in the ESM).
